---
title: "Reproducible Research: Peer Assessment 1"
author: Daniel Escasa
output: 
  html_document:
    number_sections: TRUE
    keep_md: true
editor_options: 
  markdown: 
    wrap: 72
---
```
{r setup, include=FALSE} knitr::opts_chunk$set(echo = TRUE)
```
# Introduction {.unnumbered}

*Excerpts from the [original
repo](https://github.com/rdpeng/RepData_PeerAssessment1) from which this
was forked*

It is now possible to collect a large amount of data about personal
movement using activity monitoring devices such as a
[Fitbit](http://www.fitbit.com), [Nike
Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or [Jawbone
Up](https://jawbone.com/up). These type of devices are part of the
"quantified self" movement -- a group of enthusiasts who take
measurements about themselves regularly to improve their health, to find
patterns in their behavior, or because they are tech geeks. But these
data remain under-utilized both because the raw data are hard to obtain
and there is a lack of statistical methods and software for processing
and interpreting the data.

This project makes use of data from a personal activity monitoring
device which collects data at five-minute intervals through out the day.

The data consists of two months of data from an anonymous individual
collected during the months of October and November, 2012 and include
the number of steps taken in five-minute intervals each day.

+---------------+-----------------------------------------------------+
| Variable name | Description                                         |
+===============+=====================================================+
| steps         | Number of steps taking in a five-minute interval    |
|               | (missing values are coded as \`NA\`)                |
+---------------+-----------------------------------------------------+
| date          | The date on which the measurement was taken, in     |
|               | YYYY-MM-DD format                                   |
+---------------+-----------------------------------------------------+
| interval      | Identifier for the five-minute interval in which    |
|               | measurement was taken                               |
+---------------+-----------------------------------------------------+

The following are the questions to be addressed:

-   What is mean total number of steps taken per day?

-   What is the average daily activity pattern?

-   Are there differences in activity patterns between weekdays and
    weekends?

# Boring admin stuff

## Install ggplot2 if necessary

```{r echo = TRUE}
if (!require("ggplot2")) {
    message("Installing ggplot2")
    install.packages("ggplot2")
}
```

## Download the zipped dataset if necessary

```{r echo = TRUE}
if (!file.exists("./activity.zip")) {
          message("Downloading dataset")
          download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", 
                        destfile = "./activity.zip",  
                        method = "internal", 
                        mode = "wb")
     }
```

## Unzip the dataset if necessary

```{r echo = TRUE}
if (!file.exists(".activity.csv")) {
          message("Extracting dataset")
          unzip("./activity.zip", 
                overwrite = FALSE)
     }
```

## Read in the data
```{r echo = TRUE}
# Shouldn't need error checking here
activity <- read.csv("activity.csv")
```

# Now let's get cooking

## What is mean total number of steps taken per day?

```{r echo = TRUE}
# compute the total number of steps per day
totalDaySteps <- with(data = activity,
                      aggregate(steps,
                                by  = list(date),
                                FUN = sum))
colnames(totalDaySteps) <- c("date", "steps")

# compute the mean and median number of steps per day
meanDaySteps <- with(data = activity,
                     aggregate(steps,
                               by  = list(date),
                               FUN = mean))
colnames(meanDaySteps) <- c("date", "steps")

medianDaySteps <- with(data = activity,
                       aggregate(steps,
                                 by  = list(date),
                                 FUN = median))
colnames(medianDaySteps) <- c("date", "steps")
```

## What is the average daily activity pattern?

## Imputing missing values

## Are there differences in activity patterns between weekdays and weekends?
